Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-03-05 paper/manuscript.tex
Restart writing a bit ..  On branch sboresch
 Changes to be committed:
	modified:   manuscript.tex
	modified:   sb.bib

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a quite similar approach \cite{Lee_2023}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

Dual coordinate approaches with chemical topologies encompass a third approach, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space.

@@

Some additional terms have also been employed to talk about these different intermediate pathways . Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.

>>>>

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

Dual coordinate approaches with chemical topologies encompass a third approach, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

%Some additional terms have also been employed to talk about these different intermediate pathways . Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.

Unfortunately, during the past years the use of the terms single and dual topology has proliferated, often with quite different meaning from the original intentions. The above presentation makes clear that the primary difference between approaches is the use of separate coordinates for each end state (chemical topology) or the use of an alchemical chimera with a single set of coordinates (alchemical topology). We therefore suggest the following classification
\begin{itemize}
\item Single coordinate (SC) --- Alchemical topology
  \begin{itemize}
  \item Single topology (SCST)
    \item Dual topology (SCDT)
  \end{itemize}
  \item Dual coordinate --- Chemical topology (DC)
  \end{itemize}
For completeness, we list some additional terminology that has been used: Some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though in our view these calculations should be classified as single coordinate single topology \sbnote{Check whether this is correct!}. %this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
Ries et al.\ suggested terminology \cite{Ries_2022} in which single topology corresponds to SCST, hybrid topology to SCDT, and dual topology to DC.
